---
title: "Binomial and Count Outcomes in `SteppedPower`"
author:
- name: Philipp Mildenberger
  affiliation: Institute of Medical Biostatistics, Epidemiology and Informatics ([IMBEI, Mainz](https://www.unimedizin-mainz.de/imbei/imbei/welcome-page/))
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Binomial and Count Outcomes in `SteppedPower`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  warning = FALSE
)
library(knitr)
library(SteppedPower)
library(Matrix)
library(plotly)
assign("knit_print.plotly", function(x, ...) {
  x <- plotly::partial_bundle(x)
  NextMethod()
}, envir = .GlobalEnv)

# A function for captioning and referencing images
fig <- local({
    i <- 0
    ref <- list()
    list(
        cap=function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- i
            paste("Figure ", i, ": ", text, sep="")
        },
        ref=function(refName) {
            paste("(Fig.", ref[[refName]],")")
        })
})
```

<!-- ```{css, echo=FALSE}
.columns {
  display: flex;
  gap: 2em;
}
.column {
  flex: 1;
}
``` -->

# Methods and Notation

For a binomial outcome we keep the model from the [Getting Started](Getting_Started.html) vignette
[@hussey2007design; @li2020mixed], but with an **identity link**:

$$y_{ijk}= T_{ij} \theta + c_i + \mu_j + e_{ijk}$$

with $T_{ij}$ the treatment status, $\theta$ the treatment effect, and
$c_i \sim N(0,\tau^2)$ a random cluster effect. Because the link is the
identity, the GLS machinery of `glsPower()` applies directly. The key
difference to the Gaussian case is that the residual variance is tied to
the mean:

$$\sigma^2_{ij} = p_{ij}(1-p_{ij}), \qquad p_{ij} = \mu_0 + T_{ij}\,\theta .$$

$\sigma^2$ is not a free parameter, but is rather specified by the cell prevalences $p_{ij}$. Passing `sigma` for a binomial family
triggers a warning and is ignored.

By default, `mu0` and `mu1` are the prevalence **conditional on the random
effects being zero**. With `marginal_mu = TRUE` they are read as **marginal**
prevalences and back-transformed via numerical integration. This only matters
when $\tau$ is non-negligible.

> Note that `glsPower()` derives $\sigma^2$ cell-by-cell and not just once using the mean prevalence $p_{\text{mid}}=(\mu_0+\mu_1)/2$. This may cause differences to other implementations.


# Covariance Structure Specifications

`SteppedPower` offers two ways to specify the within-cluster correlation, as described in the `Getting Started` vignette. For linear outcomes, the two routes are equivalent. For binomial outcomes, they differ because the residual variance changes accross cells.

## Random effects

In this approach, the standard deviations of the random terms are specified directly: `tau`
(cluster intercept), `gamma` (time effect), `psi` (subject intercept), `eta`
(random treatment effect), and `AR` (AR(1) decay). This route is required when
you need `eta`, `rho` (intercept-slope correlation), or `AR`.

## $\alpha_0$, $\alpha_1$, $\alpha_2$

The second implemented approach follows notation from [@li2018sample]. Here the
correlation structure is described by three scalars, $\alpha_0$, $\alpha_1$,
and $\alpha_2$, which provide an alternative parameterization of the ICC, CAC,
and IAC correlations (see the [Getting Started](Getting_Started.html) vignette).

Given a marginal variance $\sigma^2_{\text{marg}}$, these translate into
random effects via

$$
\begin{align*}
  \tau^2            &:= \sigma^2_{\text{marg}} \cdot \alpha_1 \\
  \gamma^2          &:= \sigma^2_{\text{marg}} \cdot (\alpha_0 - \alpha_1) \\
  \psi^2            &:= \sigma^2_{\text{marg}} \cdot (\alpha_2 - \alpha_1) \\
  \sigma^2_{\text{res}} &:= \sigma^2_{\text{marg}} \cdot (1 - \alpha_0 - \alpha_2 + \alpha_1)
\end{align*}
$$

If `alpha_0_1_2` of length 2 is passed, a cross-sectional design is assumed and
$\alpha_2$ is set equal to $\alpha_1$. As with the ICC/CAC/IAC
parameterization, $\boldsymbol{\alpha}$ cannot specify a
random treatment effect.

# Example Usage

:::::: {.columns}
::: {.column width="40%"}

Suppose we are planning a stepped wedge trial with four sequences
with one cluster each, and we expect a baseline prevalence of
$p=0.2$ and a treatment effect of $\theta=0.15$. We want to target
an ICC of $0.2$.

:::
::: {.column width="60%"}

```{r design-plot, fig.cap=fig$cap("design","Design matrix for 4 sequences with 1 cluster each. 5 periods."), fig.width=2, fig.height=1, echo=FALSE}
dsn_bin <- construct_DesMat(Cl = rep(1, 4), N = 50)
plot(dsn_bin)
```
:::
::::::


The two calls below target the same ICC of $0.2$ but yield different power.
With $p=0.2$ and $\theta=0.15$, the binomial variance ranges from $0.16$
(control) to $0.2275$ (treatment). Under random effects, `tau` is a fixed scalar, usually relative to the midpoint prevalence $p_{\text{mid}}=0.275$. Hence, the implied ICC varies from $0.31$ to $0.22$ across cells. Under `alpha_0_1_2`, `tau` is recalculated cell by cell and the ICC stays at $0.2$ throughout.

```{r}
p <- 0.2; effect <- 0.15; icc <- 0.2
p_mid <- p + effect * 0.5

DM <- construct_DesMat(Cl = rep(1,4), N = 50)
pow_re <- glsPower(DesMat = DM, mu0 = p, mu1 = p + effect, 
                   tau = sqrt(p_mid * (1 - p_mid) * icc / (1 - icc)),
                   family = "binomial", verbose = 2)

pow_al <- glsPower(DesMat = DM, mu0 = p, mu1 = p + effect,
                   alpha_0_1_2 = c(icc, icc),
                   family = "binomial", verbose = 2)

c(power_randomEffects = pow_re$power,
  power_alpha012      = pow_al$power)
```

```{r, fig.cap=fig$cap("covmat","Covariance matrix under random effects (left) and alpha_0_1_2 (right), both targeting ICC = 0.2.")}
plotly::subplot(
  plot(pow_re, which = 4, show_colorbar = FALSE)$CMplot,
  plot(pow_al, which = 4, show_colorbar = FALSE)$CMplot,
  nrows = 1, margin = 0.05)
```

>CAVE: $\sigma^2_{ij}=p_{ij}(1-p_{ij})$ is cell-specific -- $\sigma^2$ moves
with $p$, so the *same* $\tau$ maps to a different ICC in control vs.
intervention periods. The alpha parameterization makes this explicit (alphas
are ratios to $\sigma^2_{\text{marg}}$); random effects keeps $\tau$ fixed and
lets the implied ICC drift.

The direction reverses when $p$ and $p_1$ are symmetric around $0.5$
(here $p=0.25$, $p_1=0.75$), so both arms share the same variance
$\sigma^2=0.1875$ and there is no ICC drift. The difference now comes from the
anchoring: random effects derives $\tau$ from $p_{\text{mid}}=0.5$, whose
variance $0.25$ exceeds the actual cell variance $0.1875$, inflating $\tau$
and lowering power.



## Differences in the Variance estimator $\text{Var}(\hat\theta)$ the $(p_0, p_1)$ Plane

The contour plots below shows the difference in treatment effect variance estimators across the $(p_0, p_1)$ plane for 
the example setting above. The diagonal ($p_0=p_1$, no
treatment effect) is zero by construction. 

```{r variance computation, echo=FALSE}
# data-raw/binom_outcomes_grid.R for the computation.
# vals <- readRDS("inst/vignettes/BinomOutcomes_vals.rds")
vals <- readRDS(system.file("vignettes", 
                            "BinomOutcomes_vals.rds",
                            package = "SteppedPower"))

p_vals <- unique(vals$p0)
reMat  <- matrix(vals[, "re"],  length(p_vals), length(p_vals))
alMat  <- matrix(vals[, "al"],  length(p_vals), length(p_vals))
linMat <- matrix(vals[, "lin"], length(p_vals), length(p_vals))
```

### Cell-by-cell calculation of $\sigma^2$ versus midpoint method

```{r contour_diff, fig.width=6.5, fig.height=5.5, echo=FALSE, fig.cap=fig$cap("contour_midpoint","Relative difference of variance estimators. ICC=0.2, N=50, 4 sequences. blue: midpoint method has higher variance than cell-by-cell method.")}
diff_mat <- (reMat - linMat)/linMat * 100
maxabs <- max(abs(diff_mat), na.rm = TRUE)
plot_ly(
  x = p_vals, y = p_vals, z = diff_mat,
  type = "contour",
  colorscale = list(list(0,   "steelblue"),
               list(0.5, "white"),
               list(1,   "firebrick") ),
  zmin = -maxabs, zmax = maxabs,
  contours = list(coloring = "fill", showlabels = TRUE),
  line  = list(smoothing = 1.3) ) |>
    add_trace(x = range(p_vals), y = range(p_vals), type = "scatter",
                mode = "lines", line = list(color = "gray", dash = "dash"),
                showlegend = FALSE, hoverinfo = "none") |>
    layout( xaxis = list(title = "p0"),
            yaxis = list(title = "p1") )
```


### Random effects versus alpha notation

```{r contour_diff2, fig.width=6.5, fig.height=5.5, echo=FALSE, fig.cap=fig$cap("contour_alpha","Relative difference of variance estimators. ICC=0.2, N=50, 4 sequences. blue: alpha method has higher variance than random effects method.")}
diff_mat <- (reMat - alMat)/alMat * 100
maxabs <- max(abs(diff_mat), na.rm = TRUE)
plot_ly(
  x = p_vals, y = p_vals, z = diff_mat,
  type = "contour",
  colorscale = list(list(0,   "steelblue"),
               list(0.5, "white"),
               list(1,   "firebrick") ),
  zmin = -maxabs, zmax = maxabs,
  contours = list(coloring = "fill", showlabels = TRUE),
  line  = list(smoothing = 1.3)) |>
    add_trace(x = range(p_vals), y = range(p_vals), type = "scatter",
                   mode = "lines", line = list(color = "gray", dash = "dash"),
                   showlegend = FALSE, hoverinfo = "none") |>
    layout( xaxis = list(title = "p0"),
            yaxis = list(title = "p1") )
```


# Count (Poisson) Outcomes

Count outcomes are handled very similary in  `SteppedPower`. It uses a Poisson distribution with an **identity link**:

$$y_{ijk}= T_{ij} \theta + c_i + \mu_j + e_{ijk}$$

As with the binomial case, the residual variance is tied to the mean via $\sigma^2 = \mu$:

$$\sigma^2_{ij} = \mu_{ij} = \mu_0 + T_{ij}\,\theta .$$
Again, $\sigma^2$ is derived cell-by-cell. The same two parametrisations of the covariance structure are available with the same implications. 

# References {-}
