---
title: "Incomplete (Stepped Wedge) Designs with `SteppedPower`"
author:
- name: Philipp Mildenberger^[pmildenb@uni-mainz.de]
  affiliation: Institute of Medical Biostatistics, Epidemiology and Informatics ([IMBEI, Mainz](https://www.unimedizin-mainz.de/imbei/imbei/welcome-page/))
- name: Federico Marini^[marinif@uni-mainz.de]
  affiliation: Institute of Medical Biostatistics, Epidemiology and Informatics ([IMBEI, Mainz](https://www.unimedizin-mainz.de/imbei/imbei/welcome-page/))
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Incomplete Stepped Wedge Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SteppedPower)
```

# Incomplete Designs

In general, a study design is referred to as incomplete if not all clusters are
observed at every time period [@hemming2015stepped]. 

Suppose you do not plan to observe all clusters over the whole study period.
Rather, clusters that switch early to the intervention 
are not observed until the end.
Analogous, observation starts later in clusters 
that switch towards the end of the study.


# Incomplete Designs in `SteppedPower`

There are essentially three ways to define cluster periods without observation. 

* The first - and generally preferred - option is to use the `incomplete` argument. 
Input can be either a scalar or a matrix of dimension clusters$\cdot$timepoints 
or sequences$\cdot$timepoints:
  * A scalar is interpreted as the number of observed periods before and 
after the switch from control to intervention in each cluster
  * A matrix must contain `1`s for cluster cells that are observed and `0` or `NA`s for
  cluster cells that are not observed.
* Insert `NA`s into an explicitly defined treatment matrix, easiest done with the argument `trtmatrix=`.
* Insert `NA`s into the vector for delayed treatment start `trtDelay=`.

`glsPower()` calls the function `construct_DesMat()` to construct 
the design matrix with the relevant arguments. 
All the above options can be used in the main wrapper function, but
the examples below focus on `construct_DesMat()` directly.

> `SteppedPower` stores information about (un)observed cluster cells separately from
the treatment allocation. This is done for more consistency in the code as the 
indices in the covariance and design matrices is 

# Examples

## 1
If for example the a stepped wedge study consists of eight clusters 
in four sequences (i.e. five timepoints), 
and we observe two timepoints before and after the switch, then we receive

```{r}
Dsn1.1 <- construct_DesMat(Cl=rep(2,4), incomplete=2)
```

A slightly more tedious, but more flexible way is to define a matrix 
where each row corresponds to either a cluster or a wave of clusters 
and each column corresponds to a timepoint. 
If a cluster is not observed at a specific timepoint, 
set the value in the corresponding cell to `0`. 
For the example above, such a matrix would look like this:

```{r}
TM  <- toeplitz(c(1,1,0,0))
incompleteMat1 <- cbind(TM[,1:2],rep(1,4),TM[,3:4])
incompleteMat2 <- incompleteMat1[rep(1:4,each=2),]
```

A matrix where each row represents a wave of clusters

```{r, echo=FALSE}
suppressWarnings(knitr::kable(incompleteMat1))
```

or each row represents a cluster

```{r, echo=FALSE}
suppressWarnings(knitr::kable(incompleteMat2))
```

Now all that's left to do is to plug that into the function and we receive the 
same design matrix

```{r}
Dsn1.2 <- construct_DesMat(Cl=rep(2,4), incomplete=incompleteMat1)
Dsn1.3 <- construct_DesMat(Cl=rep(2,4), incomplete=incompleteMat2)

all.equal(Dsn1.1,Dsn1.2)
all.equal(Dsn1.1,Dsn1.3)
```


> The argument `incomplete` with matrix input works also for other design types, 
but makes (supposedly) most sense in the context of stepped wedge designs


## 2

Now suppose we want to use a SWD to investigate the intervention effects after at least one month,  
i.e. cluster periods directly after the switch to intervention conditions are not observed. 
That leads to an incomplete design that is easiest modelled with `trtDelay=`


```{r}
Dsn2 <- construct_DesMat(Cl=rep(2,4), trtDelay = c(NA) )
Dsn2
```

## 3

The above arguments can also be combined, e.g.
```{r}
Dsn3 <- construct_DesMat(Cl=rep(2,4), incomplete=2, trtDelay=c(NA) )
Dsn3
```

